A Proposal for Linguistic Similarity Datasets Based on Commonality Lists

### 2016-06-17 paper.tex

Since there is no way of deciding upfront whether there is a similarity relation between two words, the data collection procedure needs to test for both: relation existence and its strength. Numerical values, as has been shown in the introduction, do not fit this role, because of ambiguity. One way of avoiding the issue is to avoid asking humans for numerical similarity judgements, but instead to ask them to list commonalities and differences between the objects. As one might expect, similarity scores correlate with the number of listed commonalities \cite{markman1991commonalities,Markman1996,medin1993respects}. For incomparable pairs, the commonality list should be empty, but the differences will enumerate properties that belong to one entity, but not another \cite{markman1991commonalities,medin1993respects}. Verbally produced features (norms) for empirically derived conceptual representation of \newcite{McRae2005} is a good example of what and how the data should be collected. But in contrast to \newcite{McRae2005}---where explicit comparison of concepts was avoided---participants should be asked to produce commonalities as part of similarity comparison.

@@

So far, we have proposed a similarity judgement collection method that is robust to incomparable pairings. It also naturally gives rise to categories, because the absence of a relation between two entries means the absence of a common category. It still needs to be decided what words to include to the dataset.

@@

Given the dynamic nature of score collection, we suggest setting up \emph{a game with a purpose} (see \newcite{VenhuizenBasileEvangBos2013IWCS} for an example) where players are rewarded for contributing their commonality lists.

@@

Similarity evaluation needs to focus on how well a model is able to recover human similarity intuitions expressed as groupings, possibly around their focal points. We propose to treat it as a soft multi-class clustering problem \cite{White:2015:WSE:2838931.2838932}, where two entities belong to a same class if there is a similarity judgement for them (e.g.~\textit{apple} and \textit{banana} are similar because they are \textit{fruits}) and the strength is proportional to the number of such judgements, so we could express that \textit{apple} is more a \textit{fruit} than a \textit{company}.

@@

To be able to measure similarity of phrases and sentences in the proposed framework, we need to be able to identify what could serve as commonalities between them. So what are they? First of all, words, sentences and other constructions draw attention to \emph{states of affairs} around us. 
% This is both pragmatically and semantically speaking the case.

@@

Some sentences are similar to others with respect to the functions they perform. As \newcite{tomasello2009cultural} points out, speakers of English can make sense of phrases like \textit{X floosed Y the Z} and \textit{X was floosed by Y}. This is due to their similarity to sentences such as \textit{John gave Mary the book} and \textit{Mary was kissed by John} respectively. Thus, \textit{X floosed Y the Z} is clearly a \emph{transfer of possession} or \emph{dative} \cite{bresnan2007predicting}.

The amount in which sentences are similar, at least to a certain extent, corresponds to the function of a given sentence. \newcite{tomasello1998return} points out that sentence-level constructions show \emph{prototype effects} similar to those discussed above for lexical systems (e.g.~colours). Consider the following sentences:
\begin{compactitem}
    \item \textit{John gave Mary the book.} is a example of an \emph{Agent Causes Transfer} construction. These usually are build around words such as \textit{give, pass, hand, toss, bring, etc.}
      \item \textit{John promised Mary the book.} is a example of an \emph{Conditional transfer} construction. These usually are build around words such as \textit{promise, guarantee, owe, etc.}
    %   \item \textit{``John denied Mary the book.''} is a example of an \emph{Agent Prevents Transfer} construction. These usually are build around words such as \textit{refuse, deny, etc.}
    %   \item \textit{``John left Mary the book''} is a example of an \emph{Future Transfer} construction. These usually are build around words such as \textit{bequeath, leave, reserve, grant, etc.}
    %   \item \textit{``John allowed Mary one book''} is a example of an \emph{Enabling Conditions for Transfer} construction. These usually are build around words such as \textit{permit, allow, etc.}
    %   \item \textit{``John wrote Mary a book''} is a example of an \emph{Intended Transfer} construction. These usually are build around words such as \textit{bake, build, make, get, grab, win, earn, etc.}
\end{compactitem}

As soon as one has such a prototype network, one can actually decide sentence similarity as one can say with respect to what sentences and utterances are similar.

@@

However, prototype categories work on the semantic-grammatical level, and might be handled by \emph{similarity in context}: a noun phrase can be similar to a noun as in \textit{female lion} and \textit{lioness}, and to another noun phrases as in \textit{yellow car} and \textit{cheap taxi}. The same similarity principle can be applied to phrases as to words. In this case, similarity is measured in context, but is still comparison of the phrases' head words which meaning is modified by arguments they appear with \cite{Kintsch2001173,mitchell-lapata:2008:ACLMain,mitchell2010composition,Dinu:2010:MDS:1870658.1870771,Baroni2010nouns,thater-furstenau-pinkal:2011:IJCNLP-2011,Seaghdha:2011:PMS:2145432.2145545}. With verbs this idea can be applied to compare transitive verbs with intransitive. For example, \textit{to cycle} is similar to \textit{to ride a bicycle}.

Sentential similarity might be treated as the similarity of the heads in the contexts. That is, the similarity between \textit{sees} and \textit{notices} in \textit{John \textbf{sees} Mary} and \textit{John \textbf{notices} a woman}. This approach abstracts away grammatical difference between the sentences and concentrates on semantics and fits the proposed model as the respect for the head, which is a lexical entity, has to be found \cite{corbett1993heads}.

@@

But still, what about pragmatics? As \newcite{Steels2008} points out, sentences and words draw attention and do not always directly point or refer to entities and actions in the world. For example, he points to the fact that if a person asks another person to \textit{pass the wine} they are actually asking for the \textit{bottle}. The speaker just attracts attention to an object of perception in a given situation.

@@

There are several ways in which a sentence can both be \emph{grammaticalised} and \emph{lexicalised}. For example, \textit{No} and \textit{I've seen John eating them} are similar sentences because they lexicalise the same answer to the question \textit{Do we have cookies?} More generally, this gives rise to dialogue act tags: another way of utterance categorisation, refer to the work of \newcite{kalchbrenner-blunsom:2013:CVSC} and \newcite{milajevs-purver:2014:CVSC}.

Thus, questions, which the sentences answer, are valid respects for similarity explanation, as well as entailment, paraphrase \cite{White:2015:WSE:2838931.2838932} or spatial categories \cite{ritter-EtAl:2015:*SEM2015}. This also motivates the approach of treating sentences on their own and encoding the meaning of a sentence into a vector in such a way that similar sentences are clustered together \cite{DBLP:journals/corr/abs-1003-4394,baroni2014frege,Socher:2012:SCT:2390948.2391084,wieting2015paraphrase,DBLP:journals/corr/HillCK16}.

@@

If one conceptualises sentence similarity with respect to a discourse, then one might ask how different sentences fit in to such a discourse. \newcite{Griffiths2015a} tried to construct two versions of the same dialogue using a bottom-up method. They deconstructed a certain dialogue in a given domain---a receptionist scenario---into \textit{greetings}, \textit{directions} and \textit{farewells}. They used a small custom made corpus for this purpose and created the two dialogues by having people rate the individual utterances by friendliness. The resulting two dialogues were surprisingly uneven. The dialogue was supposed to give instructions to a certain location within a building. The ``friendly version'' was very elaborated and consisted of several sentences:
\enumsentence{
The questionnaire is located in room
Q2-102. That is on the second floor. If
you turn to your right and walk down
the hallway. At the end of the floor you
will find the stairs. Just walk up the
stairs to the top floor and go through
the fire door. The room is then straight
ahead.}
The sentence which served the same purpose in the neutral version was a fairly simple sentence:
\enumsentence{
The questionnaire is located in Q2-102.
}
Often the same functions in dialogue can be performed by as little as one word or several phrases or even a complete story.

@@

\newcite{Steels2010a} introduces the idea of language sub-systems and language strategies. A language subsystem are the means of expressing certain related or similar meanings. Examples of such sub-systems include:
\begin{compactitem}
    \item Lexical systems which express colours.
    \item Morphological devices to encode tenses.
    \item Usage of word order to express relations between agent and patient.
\end{compactitem}  
The later is an illustration of a language strategy. In English agent-patient relations are mainly encoded by syntax whereas German would use intonation and a combination or articles and case to convey the same information. Russian, in contrast, will use morphological devices for the same purpose.

Above we identified seven criteria by which sentence similarity can be compared.

@@

In the future, a multi-category similarity dataset should be build that allows evaluation of vector space models of meaning by not only measuring proximity between the points, but also their arrangement to clusters. The same ideas can be used to build phrase- and sentence-level datasets.

@@

>>>>

Since there is no way of deciding upfront whether there is a similarity relation between two words, the data collection procedure needs to test for both: relation existence and its strength. Numerical values, as has been shown in the introduction, do not fit this role due to ambiguity. One way to avoid the issue is to avoid asking humans for numerical similarity judgements, but instead to ask them to list commonalities and differences between the objects. As one might expect, similarity scores correlate with the number of listed commonalities \cite{markman1991commonalities,Markman1996,medin1993respects}. For incomparable pairs, the commonality list should be empty, but the differences will enumerate properties that belong to one entity, but not to another \cite{markman1991commonalities,medin1993respects}.

Verbally produced features (norms) for empirically derived conceptual representation of \newcite{McRae2005} is a good example of what and how the data should be collected. But in contrast to \newcite{McRae2005}---where explicit comparison of concepts was avoided---participants should be asked to produce commonalities as part of similarity comparison.

@@

So far, we have proposed a similarity judgement collection method that is robust to incomparable pairings. It also naturally gives rise to categories, because the absence of a relation between two entries means the absence of a common category. It still needs to be decided which words to include in the dataset.

@@

Given the dynamic nature of score collection, we suggest setting up \emph{a game with a purpose} (see \newcite{VenhuizenBasileEvangBos2013IWCS} an example) where players are rewarded for contributing their commonality lists. Another option would be to crowdsource the human judgements \cite{doi:10.1080/17470218.2015.1051065}.

@@

Similarity evaluation needs to focus on how well a model is able to recover human similarity intuitions expressed as groupings, possibly around their focal points. We propose to treat it as a soft multi-class clustering problem \cite{White:2015:WSE:2838931.2838932}, where two entities belong to the same class if there is a similarity judgement for them (e.g.~\textit{apple} and \textit{banana} are similar because they are \textit{fruits}) and the strength is proportional to the number of such judgements, so we could express that \textit{apple} is more a \textit{fruit} than it is a \textit{company}.

@@

To be able to measure the similarity of phrases and sentences in the proposed framework, we need to be able to identify what could serve as commonalities between them. So what are they? First of all, words, sentences and other constructions draw attention to \emph{states of affairs} around us. Also, sentences are similar to others with respect to the functions they perform \cite[p. 288]{Winograd1983}. 
% This is both pragmatically and semantically speaking the case.

@@

As \newcite{tomasello2009cultural} points out, speakers of English can make sense of phrases like \textit{X floosed Y the Z} and \textit{X was floosed by Y}. This is due to their similarity to sentences such as \textit{John gave Mary the book} and \textit{Mary was kissed by John} respectively. Thus, \textit{X floosed Y the Z} is clearly a \emph{transfer of possession} or \emph{dative} \cite{bresnan2007predicting}.

The amount in which sentences are similar, at least to a certain extent, corresponds to the function of a given sentence (the ideational function \cite[p. 288]{Winograd1983} especially). \newcite{tomasello1998return} points out that sentence-level constructions show \emph{prototype effects} similar to those discussed above for lexical systems (e.g.~colours). Consider the following sentences:
\begin{compactitem}
    \item \textit{John gave Mary the book.} is a example of an \emph{Agent Causes Transfer} construction. These usually are build around words such as \textit{give, pass, hand, toss, bring, etc.}
      \item \textit{John promised Mary the book.} is a example of an \emph{Conditional transfer} construction. These usually are build around words such as \textit{promise, guarantee, owe, etc.}
    %   \item \textit{``John denied Mary the book.''} is a example of an \emph{Agent Prevents Transfer} construction. These usually are build around words such as \textit{refuse, deny, etc.}
    %   \item \textit{``John left Mary the book''} is a example of an \emph{Future Transfer} construction. These usually are build around words such as \textit{bequeath, leave, reserve, grant, etc.}
    %   \item \textit{``John allowed Mary one book''} is a example of an \emph{Enabling Conditions for Transfer} construction. These usually are build around words such as \textit{permit, allow, etc.}
    %   \item \textit{``John wrote Mary a book''} is a example of an \emph{Intended Transfer} construction. These usually are build around words such as \textit{bake, build, make, get, grab, win, earn, etc.}
\end{compactitem}

As soon as one has such a prototype network, one can actually decide sentence similarity as one can say with respect to what prototypes sentences and utterances are similar. In this case, a common sentence prototype serves the same role as commonality between words.

@@

However, prototype categories work on the semantic-grammatical level, and might be handled by \emph{similarity in context}: a noun phrase can be similar to a noun as in \textit{female lion} and \textit{lioness}, and to another noun phrases as in \textit{yellow car} and \textit{cheap taxi}. The same similarity principle can be applied to phrases as to words. In this case, similarity is measured in context, but it is still a comparison of the phrases' head words of which meaning is modified by arguments they appear with \cite{Kintsch2001173,mitchell-lapata:2008:ACLMain,mitchell2010composition,Dinu:2010:MDS:1870658.1870771,Baroni2010nouns,thater-furstenau-pinkal:2011:IJCNLP-2011,Seaghdha:2011:PMS:2145432.2145545}. With verbs this idea can be applied to compare transitive verbs with intransitive. For example, \textit{to cycle} is similar to \textit{to ride a bicycle}.

Sentential similarity might be treated as the similarity of the heads in the contexts. That is, the similarity between \textit{sees} and \textit{notices} in \textit{John \textbf{sees} Mary} and \textit{John \textbf{notices} a woman}. This approach abstracts away from grammatical differences between the sentences and concentrates on semantics and fits the proposed model as the respect for the head, which is a lexical entity, has to be found \cite{corbett1993heads}.

@@

But still, what about pragmatics? As \newcite{Steels2008} points out, sentences and words direct attention and do not always directly point or refer to entities and actions in the world. For example, he points to the fact that if a person asks another person to \textit{pass the wine} they are actually asking for the \textit{bottle}. The speaker just attracts attention to an object of perception in a given situation.

@@

There are several ways in which a sentence can both be \emph{grammaticalised} and \emph{lexicalised}. For example, \textit{No} and \textit{I've seen John eating them} are similar sentences because they lexicalise the same answer to the question \textit{Do we have cookies?} More generally, this gives rise to dialogue act tags: for another way of utterance categorisation, refer to the work of \newcite{kalchbrenner-blunsom:2013:CVSC} and \newcite{milajevs-purver:2014:CVSC}.

Thus, questions which the sentences answer, are valid respects for similarity explanation, as well as entailment, paraphrase \cite{White:2015:WSE:2838931.2838932} or spatial categories \cite{ritter-EtAl:2015:*SEM2015}. This also motivates the approach of treating sentences on their own and encoding the meaning of a sentence into a vector in such a way that similar sentences are clustered together \cite{DBLP:journals/corr/abs-1003-4394,baroni2014frege,Socher:2012:SCT:2390948.2391084,wieting2015paraphrase,DBLP:journals/corr/HillCK16}.

@@

If one conceptualises sentence similarity with respect to a discourse, then one might ask how different sentences fit in to such a discourse. \newcite{Griffiths2015a} tried to construct two versions of the same dialogue using a bottom-up method. They deconstructed a certain dialogue in a given domain---a receptionist scenario---into \textit{greetings}, \textit{directions} and \textit{farewells}. They used a small custom made corpus for this purpose and created the two dialogues by having people rate the individual utterances by friendliness. The resulting two dialogues were surprisingly uneven. The dialogue was supposed to give instructions to a certain location within a building. The ``friendly version'' was very elaborated and consisted of several sentences:
\enumsentence{
The questionnaire is located in room
Q2-102. That is on the second floor. If
you turn to your right and walk down
the hallway. At the end of the floor you
will find the stairs. Just walk up the
stairs to the top floor and go through
the fire door. The room is then straight
ahead.}
The sentence which served the same purpose in the ``neutral version'' was a fairly simple sentence:
\enumsentence{
The questionnaire is located in Q2-102.
}
Often the same function of a given sentence in a dialogue can be performed by as little as one word or several phrases or a different sentence or even a complete story.

@@

\newcite{Steels2010a} introduces the idea of language sub-systems and language strategies. A language subsystem are the means of expressing certain related or similar meanings. Examples of such sub-systems include:
\begin{compactitem}
    \item Lexical systems which express colours.
    \item Morphological devices to encode tenses.
    \item Usage of word order to express relations between agent and patient.
\end{compactitem}  
The later is an illustration of a language strategy. In English agent-patient relations are mainly encoded by syntax whereas German would use intonation and a combination or articles and case to convey the same information. Russian, in contrast, will use morphological devices for the same purpose. Hence, for some purposes the entities which are similar may not be of clearly delineated categories such as ``word'' or ``sentence'' but may be of chunks of language which belong to the same sub-system.

Above we identified seven criteria by which sentence similarity can be compared. The instructions for the sentence similarity judgement tasks may incorporate the criteria as hints for human participants during data collection.

@@

In the future, a multi-category similarity dataset should be build that allow evaluation of vector space models of meaning by not only measuring proximity between the points, but also their arrangement with respect to clusters. The same ideas can be used to build phrase- and sentence-level datasets. However, we leave the exact sentence similarity criteria selection for future work in this area.

@@

\section*{Acknowledgements}

We thank the anonymous reviewers for their comments. Support from EPSRC grant EP/J002607/1 is gratefully acknowledged by Dmitrijs Milajevs.
Sascha Griffiths is supported by ConCreTe: the project ConCreTe acknowledges the financial support of the Future and Emerging Technologies (FET) programme within the Seventh Framework Programme for Research of the European Commission, under FET grant number 611733.
